---
title: "Video #7 code"
subtitle: "Overfitting"
date: "October 5, 2022"
---

<div class="ratio ratio-16x9">
<iframe src="https://www.youtube.com/embed/odGAAJDlgp8" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
</div>

```{r packages-data, warning=FALSE, message=FALSE}
library(tidyverse)
library(brms)
library(loo)
library(tidybayes)

# Plot stuff
clrs <- MetBrewer::met.brewer("Lakota", 6)
theme_set(theme_bw())

# Seed stuff
set.seed(1234)
BAYES_SEED <- 1234

data(WaffleDivorce, package = "rethinking")

WaffleDivorce <- WaffleDivorce %>% 
  mutate(across(c(Marriage, Divorce, MedianAgeMarriage), ~scale(.), .names = "{col}_scaled")) %>% 
  mutate(across(c(Marriage, Divorce, MedianAgeMarriage), ~as.numeric(scale(.)), .names = "{col}_z"))
```


# Finding outliers with PSIS and WAIC

```{r fig.width=5, fig.height=3.5, out.width="100%"}
ggplot(WaffleDivorce, aes(x = MedianAgeMarriage_z, y = Divorce_z)) +
  geom_point(aes(color = Loc %in% c("ME", "ID")), size = 2) +
  geom_text(data = filter(WaffleDivorce, Loc %in% c("ME", "ID")), 
            aes(label = Location), hjust = -0.25) +
  scale_color_manual(values = c("grey40", clrs[4]), guide = "none") +
  labs(x = "Age at marriage (standardized)", y = "Divorce rate (standardized)")
```
Run a model:

```{r age-marriage-normal, cache=TRUE, results="hide", warning=FALSE}
priors <- c(prior(normal(0, 0.2), class = Intercept),
            prior(normal(0, 0.5), class = b, coef = "Marriage_z"),
            prior(normal(0, 0.5), class = b, coef = "MedianAgeMarriage_z"),
            prior(exponential(1), class = sigma))

marriage_divorce_normal <- brm(
  bf(Divorce_z ~ Marriage_z + MedianAgeMarriage_z),
  data = WaffleDivorce,
  family = gaussian(),
  prior = priors,
  chains = 4, cores = 4, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0
)
```

Check the LOO stats. One value is fairly influential with k > 0.5, but the others are okay:

```{r}
loo(marriage_divorce_normal)
```

Which observation has the high PSIS k value?

```{r}
loo(marriage_divorce_normal) |> 
  pareto_k_ids()
```

Row 13! Which is…

```{r}
WaffleDivorce |> 
  slice(13) |> 
  select(Location)
```

Idaho.

How big is the actual k value?

```{r}
loo(marriage_divorce_normal) |> 
  pareto_k_values() |> 
  pluck(13)
```

We can embed these diagnostics into the brms object with `add_criterion()`:

```{r}
marriage_divorce_normal <- add_criterion(marriage_divorce_normal, criterion = "loo")
marriage_divorce_normal <- add_criterion(marriage_divorce_normal, criterion = "waic")
```

And that lets us access things in deeply nested lists, like the 13th Pareto k value:

```{r}
marriage_divorce_normal$criteria$loo$diagnostics$pareto_k[13]
```

Neat. Now we can plot these k values and WAIC values and recreate Figure 7.10 from the book and from 1:03:00 in lecture video 7.

```{r}
tibble(psis = marriage_divorce_normal$criteria$loo$diagnostics$pareto_k,
       p_waic = marriage_divorce_normal$criteria$waic$pointwise[, "p_waic"],
       Location = pull(WaffleDivorce, Location),
       Loc = pull(WaffleDivorce, Loc)) %>%
  ggplot(aes(x = psis, y = p_waic)) +
  geom_point(aes(color = Loc %in% c("ME", "ID")), size = 2) +
  geom_text(data = . %>% filter(Loc %in% c("ME", "ID")), 
            aes(label = Location), hjust = 1.25) +
  geom_vline(xintercept = 0.5, linetype = 32) +
  scale_color_manual(values = c("grey40", clrs[4]), guide = "none") +
  labs(x = "PSIS Pareto k", y = "WAIC penalty")
```


# Robust regression

We can do robust regression with `family = student()`, which has thicker tails and expects larger values out in the tails

```{r age-marriage-t, cache=TRUE, results="hide", warning=FALSE}
priors <- c(prior(normal(0, 0.2), class = Intercept),
            prior(normal(0, 0.5), class = b, coef = "Marriage_z"),
            prior(normal(0, 0.5), class = b, coef = "MedianAgeMarriage_z"),
            prior(exponential(1), class = sigma))

marriage_divorce_student <- brm(
  bf(Divorce_z ~ Marriage_z + MedianAgeMarriage_z,
     nu = 2),  # Tail thickness
  data = WaffleDivorce,
  family = student(),
  prior = priors,
  chains = 4, cores = 4, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0
)
```

Add penalty statistics to the model object:

```{r}
marriage_divorce_student <- add_criterion(marriage_divorce_student, criterion = c("loo", "waic"))
```

```{r}
plot_data <- tibble(psis = marriage_divorce_student$criteria$loo$diagnostics$pareto_k,
                    p_waic = marriage_divorce_student$criteria$waic$pointwise[, "p_waic"],
                    Location = pull(WaffleDivorce, Location),
                    Loc = pull(WaffleDivorce, Loc)) 
plot_data %>%
  ggplot(aes(x = psis, y = p_waic)) +
  geom_point(aes(color = Loc %in% c("ME", "ID")), size = 2) +
  geom_text(data = . %>% filter(Loc %in% c("ME", "ID")), 
            aes(label = Location), hjust = 1.25) +
  geom_vline(xintercept = 0.5, linetype = 32) +
  scale_color_manual(values = c("grey40", clrs[4]), guide = "none") +
  labs(x = "PSIS Pareto k", y = "WAIC penalty")
```

Hey hey, Idaho and Maine have much lower PSIS k values now. There are some weird observations with really high WAIC values for some reason:

```{r}
plot_data |> 
  arrange(desc(p_waic))
```

Wyoming and Utah! Why?> I don't know :(


# Compare the models

We can compare the two models' LOO statistics:

```{r}
loo_compare(marriage_divorce_normal, marriage_divorce_student, criterion = "loo")
loo_compare(marriage_divorce_normal, marriage_divorce_student, criterion = "waic")
```

The normal model has the higher ELPD score (so it's better), but the standard error is huge and makes the models indistinguishable (so it's not necessarily better)

We can also compare the posterior distributions for the effect of age on marriage. The coefficient for age in the Student-t model is more negative and more precise. Idaho was making the normal model too skeptical and too surprised; the Student-t model was less surprised by it.

```{r}
normal_coefs <- marriage_divorce_normal |> 
  spread_draws(b_MedianAgeMarriage_z) |> 
  mutate(model = "Gaussian model")

student_coefs <- marriage_divorce_student |> 
  spread_draws(b_MedianAgeMarriage_z) |> 
  mutate(model = "Student-t model")

bind_rows(normal_coefs, student_coefs) |> 
  ggplot(aes(x = b_MedianAgeMarriage_z, fill = model)) +
  stat_halfeye(slab_alpha = 0.75) +
  scale_fill_manual(values = c(clrs[6], clrs[4]))
```
