---
title: "Reading notes"
subtitle: "Extending the normal regression model"
date: "October 5, 2022"
---

[(Original chapter)](https://www.bayesrulesbook.com/chapter-11.html)

```{r warning=FALSE, message=FALSE}
library(bayesrules)
library(tidyverse)
library(brms)
library(cmdstanr)
library(rstanarm)
library(broom.mixed)
library(tidybayes)
library(ggdist)
library(patchwork)

# Plot stuff
clrs <- MetBrewer::met.brewer("Lakota", 6)
theme_set(theme_bw())

# Seed stuff
set.seed(1234)
BAYES_SEED <- 1234

data(weather_WU, package = "bayesrules")

weather_WU <- weather_WU %>% 
  select(location, windspeed9am, humidity9am, pressure9am, temp9am, temp3pm) |> 
  mutate(across(c(temp9am, temp3pm, humidity9am, windspeed9am, pressure9am), 
                ~scale(., scale = FALSE), .names = "{col}_centered")) |> 
  mutate(across(c(temp9am, temp3pm, humidity9am, windspeed9am, pressure9am), 
                ~as.numeric(scale(., scale = FALSE)), .names = "{col}_c"))

extract_attributes <- function(x) {
  attributes(x) %>%
    set_names(janitor::make_clean_names(names(.))) %>%
    as_tibble() %>%
    slice(1)
}

unscaled <- weather_WU %>%
  select(ends_with("_centered")) |> 
  summarize(across(everything(), ~extract_attributes(.))) |> 
  pivot_longer(everything()) |> 
  unnest(value) |> 
  split(~name)

# Access these things like so:
# unscaled$temp3pm_centered$scaled_center
```

# The general setup

We want to know the relationship between morning and afternoon temperatures in Wollongong and Uluru:

```{r}
ggplot(weather_WU, aes(x = temp9am, y = temp3pm)) +
  geom_point()
```

Looks linear enough. Let's make a model with vague priors (just in rstanarm and brms for now; no need for raw Stan here):

$$
\begin{aligned}
Y_i \mid \beta_0, \beta_1, \beta_2, \sigma &\stackrel{\text{ind}}{\sim} \mathcal{N}(\mu_i, \sigma^2) \\
\mu_i &= \beta_{0c} + \beta_1 X_{i1} \\
\\
\beta_{0c} &\sim \mathcal{N}(25, 5) \\
\beta_{1} &\sim \mathcal{N}(0, 2.5) \\
\sigma &\sim \operatorname{Exponential}(1)
\end{aligned}
$$

::: {.panel-tabset}

### rstanarm

```{r model-wu-rstanarm, cache=TRUE, results="hide"}
weather_simple_rstanarm <- stan_glm(
  temp3pm ~ temp9am,
  data = weather_WU,
  family = gaussian(),
  prior_intercept = normal(25, 5),
  prior = normal(0, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0
)
```

### brms

```{r model-wu-brms, cache=TRUE, results="hide"}
priors <- c(prior(normal(25, 5), class = Intercept),
            prior(normal(0, 2.5), class = b, coef = "temp9am"),
            prior(exponential(1), class = sigma))

weather_simple_brms <- brm(
  bf(temp3pm ~ temp9am),
  data = weather_WU,
  family = gaussian(),
  prior = priors,
  chains = 4, iter = 5000*2, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0
)
```

:::

We'll skip all the diagnostics for now—they cover them in the book. What really matters is how well this model fits the data, and it's not great. We've got a weird hump in the data that's not accounted for in the model. We've got to do something about that.

```{r}
pp_check(weather_simple_brms, ndraws = 25)
```

:::{.callout-note}
I went into super detail in chapters 9 and 10, with examples in rstanarm, brms, and raw Stan, and with manual calculation of core Bayesian things, like posterior predictive checks, k-fold cross validation, and a bunch of other things. 

I don't do that in this chapter, since I've been doing Bayesian regression stuff for years. Here I just do rstanarm and brms versions of everything and pre-written checks like `pp_check()`; no raw Stan here. 
:::

# 11.1: Utilizing a categorical predictor

What if we just use location to predict afternoon temperatures? It's an important factor:

```{r message=FALSE}
ggplot(weather_WU, aes(x = temp3pm, fill = location)) +
  geom_density(color = NA) +
  scale_fill_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "3 PM temperature", fill = NULL)
```

We can include it as a term in the model:

$$
\begin{aligned}
Y_i \mid \beta_0, \beta_1, \sigma &\stackrel{\text{ind}}{\sim} \mathcal{N}(\mu_i, \sigma^2) \\
\mu_i &= \beta_{0c} + \beta_1 X_{i1} \\
\\
\beta_{0c} &\sim \mathcal{N}(25, 5) \\
\beta_{1} &\sim \mathcal{N}(0, 10) \\
\sigma &\sim \operatorname{Exponential}(1)
\end{aligned}
$$

## Understanding the priors

These all look reasonable.

::: {.panel-tabset}
### rstanarm

```{r model-location-only-prior-rstanarm, cache=TRUE, results="hide"}
weather_location_only_prior_rstanarm <- stan_glm(
  temp3pm ~ location,
  data = weather_WU,
  family = gaussian,
  prior_intercept = normal(25, 5),
  prior = normal(0, 2.5, autoscale = TRUE), 
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0,
  prior_PD = TRUE
)
```

```{r}
weather_WU |> 
  add_predicted_draws(weather_location_only_prior_rstanarm, ndraws = 100) |> 
  ggplot(aes(x = .prediction, group = .draw)) +
  geom_density(size = 0.25, color = clrs[3]) +
  labs(x = "Predicted 3 PM temperature", y = "Density")
```

### brms

```{r model-location-only-prior-brms, cache=TRUE, results="hide"}
priors <- c(prior(normal(25, 5), class = Intercept),
            prior(normal(0, 10), class = b, coef = "locationWollongong"),
            prior(exponential(1), class = sigma))

weather_location_only_prior_brms <- brm(
  bf(temp3pm ~ location),
  data = weather_WU,
  family = gaussian(),
  prior = priors,
  chains = 4, iter = 5000*2, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0,
  sample_prior = "only"
)
```

```{r}
weather_WU |> 
  add_predicted_draws(weather_location_only_prior_brms, ndraws = 100) |> 
  ggplot(aes(x = .prediction, group = .draw)) +
  geom_density(size = 0.25, color = clrs[3]) +
  labs(x = "Predicted 3 PM temperature", y = "Density")
```

:::

## Simulating the posterior

::: {.panel-tabset}
### rstanarm

#### Model

```{r model-location-only-rstanarm, cache=TRUE, results="hide"}
weather_location_only_rstanarm <- stan_glm(
  temp3pm ~ location,
  data = weather_WU,
  family = gaussian,
  prior_intercept = normal(25, 5),
  prior = normal(0, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0
)
```

```{r}
tidy(weather_location_only_rstanarm, effects = c("fixed", "aux"), 
     conf.int = TRUE, conf.level = 0.8)
```

#### Coefficient posteriors

```{r}
weather_location_only_rstanarm |> 
  gather_draws(`(Intercept)`, locationWollongong, sigma) |> 
  ggplot(aes(x = .value, fill = .variable)) +
  stat_halfeye(normalize = "xy") +
  scale_fill_manual(values = c(clrs[1], clrs[3], clrs[4]), guide = "none") +
  facet_wrap(vars(.variable), scales = "free_x") +
  labs(x = "Parameter posterior") +
  theme(legend.position = "bottom")
```

#### Posterior predictive check

Ew!

```{r}
pp_check(weather_location_only_rstanarm, n = 25)
```


### brms

#### Model

```{r model-location-only-brms, cache=TRUE, results="hide"}
priors <- c(prior(normal(25, 5), class = Intercept),
            prior(normal(0, 10), class = b, coef = "locationWollongong"),
            prior(exponential(1), class = sigma))

weather_location_only_brms <- brm(
  bf(temp3pm ~ location),
  data = weather_WU,
  family = gaussian(),
  prior = priors,
  chains = 4, iter = 5000*2, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0
)
```

```{r}
tidy(weather_location_only_brms, conf.int = TRUE, conf.level = 0.8) |> 
  select(-c(effect, component, group))
```

#### Coefficient posteriors

```{r}
weather_location_only_brms |> 
  gather_draws(b_Intercept, b_locationWollongong, sigma) |> 
  ggplot(aes(x = .value, fill = .variable)) +
  stat_halfeye(normalize = "xy") +
  scale_fill_manual(values = c(clrs[1], clrs[3], clrs[4]), guide = "none") +
  facet_wrap(vars(.variable), scales = "free_x") +
  labs(x = "Parameter posterior") +
  theme(legend.position = "bottom")
```

#### Posterior predictive check

Not great!

```{r}
pp_check(weather_location_only_brms, ndraws = 25)
```

:::

## Posterior prediction

What's the predicted afternoon temperature in these two cities?

::: {.panel-tabset}
### rstanarm

```{r}
weather_location_only_rstanarm |> 
  predicted_draws(newdata = tibble(location = c("Wollongong", "Uluru"))) |> 
  ggplot(aes(x = .prediction, y = location, fill = location)) +
  stat_halfeye() +
  scale_fill_manual(values = c(clrs[2], clrs[1]), guide = "none") +
  labs(x = "Predicted 3 PM temperature", y = NULL)
```

### brms

```{r}
weather_location_only_brms |> 
  predicted_draws(newdata = tibble(location = c("Wollongong", "Uluru"))) |> 
  ggplot(aes(x = .prediction, y = location, fill = location)) +
  stat_halfeye() +
  scale_fill_manual(values = c(clrs[2], clrs[1]), guide = "none") +
  labs(x = "Predicted 3 PM temperature", y = NULL)
```

:::


# 11.2: Utilizing two predictors

There's a recognizable cluster in the data here: location. Wollongong has lower afternoon temperatures than Uluru. That makes sense—Wollongong is a coastal city near Sydney and Uluru is right in the middle of Australia in the desert.

```{r message=FALSE}
ggplot(weather_WU, aes(x = temp9am, y = temp3pm, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "9 AM temperature", y = "3 PM temperature", color = NULL)
```

Since location is binary, including it as a covariate will shift the intercept up and down

$$
\begin{aligned}
Y_i \mid \beta_0, \beta_1, \beta_2, \sigma &\stackrel{\text{ind}}{\sim} \mathcal{N}(\mu_i, \sigma^2) \\
\mu_i &= \beta_{0c} + \beta_1 X_{i1} + \beta_2 X_{i2} \\
\\
\beta_{0c} &\sim \mathcal{N}(25, 5) \\
\beta_{1} &\sim \mathcal{N}(0, 2.5) \\
\beta_{2} &\sim \mathcal{N}(0, 10) \\
\sigma &\sim \operatorname{Exponential}(1)
\end{aligned}
$$

## Understanding the priors

These all look reasonable.

::: {.panel-tabset}
### rstanarm

```{r model-location-prior-rstanarm, cache=TRUE, results="hide"}
weather_location_prior_rstanarm <- stan_glm(
  temp3pm ~ temp9am + location,
  data = weather_WU,
  family = gaussian,
  prior_intercept = normal(25, 5),
  prior = normal(0, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0,
  prior_PD = TRUE
)
```

```{r fig.width=8, fig.height=3.5}
p1 <- weather_WU |> 
  add_predicted_draws(weather_location_prior_rstanarm, ndraws = 100) |> 
  ggplot(aes(x = .prediction, group = .draw)) +
  geom_density(size = 0.25, color = clrs[3]) +
  labs(x = "Predicted 3 PM temperature", y = "Density")

# Feed add_epred_draws() brand new data instead of the original data, just for fun
prior_draws_rstanarm <- weather_WU |> 
  group_by(location) |> 
  summarize(min = min(temp9am),
            max = max(temp9am)) |> 
  mutate(temp9am = map2(min, max, ~seq(.x, .y, 1))) |> 
  unnest(temp9am) |> 
  add_epred_draws(weather_location_prior_rstanarm, ndraws = 100)

p2 <- prior_draws_rstanarm |> 
  ggplot(aes(x = temp9am, y = .epred)) +
  geom_line(aes(group = paste(location, .draw), color = location), alpha = 0.2) +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "9 AM temperature", y = "3 PM temperature", color = NULL) +
  theme(legend.position = "bottom")

p1 | p2
```

### brms

```{r model-location-prior-brms, cache=TRUE, results="hide"}
priors <- c(prior(normal(25, 5), class = Intercept),
            prior(normal(0, 2.5), class = b, coef = "temp9am_c"),
            prior(normal(0, 10), class = b, coef = "locationWollongong"),
            prior(exponential(1), class = sigma))

weather_location_prior_brms <- brm(
  bf(temp3pm ~ temp9am_c + location),
  data = weather_WU,
  family = gaussian(),
  prior = priors,
  chains = 4, iter = 5000*2, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0,
  sample_prior = "only"
)
```

```{r fig.width=8, fig.height=3.5}
p1 <- weather_WU |> 
  add_predicted_draws(weather_location_prior_brms, ndraws = 100) |> 
  ggplot(aes(x = .prediction, group = .draw)) +
  geom_density(size = 0.25, color = clrs[3]) +
  labs(x = "Predicted 3 PM temperature", y = "Density")

# Feed add_epred_draws() brand new data instead of the original data, just for fun
prior_draws_brms <- weather_WU |> 
  group_by(location) |> 
  summarize(min = min(temp9am_c),
            max = max(temp9am_c)) |> 
  mutate(temp9am_c = map2(min, max, ~seq(.x, .y, 1))) |> 
  unnest(temp9am_c) |> 
  add_epred_draws(weather_location_prior_brms, ndraws = 100) |> 
  mutate(temp9am = temp9am_c + unscaled$temp9am_centered$scaled_center)

p2 <- prior_draws_brms |> 
  ggplot(aes(x = temp9am, y = .epred)) +
  geom_line(aes(group = paste(location, .draw), color = location), alpha = 0.2) +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "9 AM temperature", y = "3 PM temperature", color = NULL) +
  theme(legend.position = "bottom")

p1 | p2
```

:::

## Simulating the posterior

::: {.panel-tabset}
### rstanarm

#### Model

```{r model-location-rstanarm, cache=TRUE, results="hide"}
weather_location_rstanarm <- stan_glm(
  temp3pm ~ temp9am + location,
  data = weather_WU,
  family = gaussian,
  prior_intercept = normal(25, 5),
  prior = normal(0, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0
)
```

```{r}
tidy(weather_location_rstanarm, effects = c("fixed", "aux"), 
     conf.int = TRUE, conf.level = 0.8)
```

#### Coefficient posteriors

```{r}
weather_location_rstanarm |> 
  gather_draws(`(Intercept)`, temp9am, locationWollongong, sigma) |> 
  ggplot(aes(x = .value, fill = .variable)) +
  stat_halfeye(normalize = "xy") +
  scale_fill_manual(values = c(clrs[1], clrs[3], clrs[4], clrs[5]), guide = "none") +
  facet_wrap(vars(.variable), scales = "free_x") +
  labs(x = "Parameter posterior") +
  theme(legend.position = "bottom")
```

#### Fitted draws

```{r}
weather_WU |> 
  add_linpred_draws(weather_location_rstanarm, ndraws = 100) |> 
  ggplot(aes(x = temp9am, y = temp3pm, color = location)) +
  geom_point(data = weather_WU, size = 0.5) +
  geom_line(aes(y = .linpred, group = paste(location, .draw)), alpha = 0.2, size = 0.5) +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "9 AM temperature", y = "3 PM temperature", color = NULL)
```

#### Posterior predictive check

Much better!

```{r}
pp_check(weather_location_rstanarm, n = 25)
```


### brms

#### Model

```{r model-location-brms, cache=TRUE, results="hide"}
priors <- c(prior(normal(25, 5), class = Intercept),
            prior(normal(0, 2.5), class = b, coef = "temp9am_c"),
            prior(normal(0, 10), class = b, coef = "locationWollongong"),
            prior(exponential(1), class = sigma))

weather_location_brms <- brm(
  bf(temp3pm ~ temp9am_c + location),
  data = weather_WU,
  family = gaussian(),
  prior = priors,
  chains = 4, iter = 5000*2, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0
)
```

```{r}
tidy(weather_location_brms, conf.int = TRUE, conf.level = 0.8) |> 
  select(-c(effect, component, group))
```

#### Coefficient posteriors

```{r}
weather_location_brms |> 
  gather_draws(b_Intercept, b_temp9am_c, b_locationWollongong, sigma) |> 
  ggplot(aes(x = .value, fill = .variable)) +
  stat_halfeye(normalize = "xy") +
  scale_fill_manual(values = c(clrs[1], clrs[3], clrs[4], clrs[5]), guide = "none") +
  facet_wrap(vars(.variable), scales = "free_x") +
  labs(x = "Parameter posterior") +
  theme(legend.position = "bottom")
```

#### Fitted draws

```{r}
weather_WU |> 
  add_linpred_draws(weather_location_brms, ndraws = 100) |> 
  ggplot(aes(x = temp9am, y = temp3pm, color = location)) +
  geom_point(data = weather_WU, size = 0.5) +
  geom_line(aes(y = .linpred, group = paste(location, .draw)), alpha = 0.2, size = 0.5) +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "9 AM temperature", y = "3 PM temperature", color = NULL)
```

#### Posterior predictive check

Much better!

```{r}
pp_check(weather_location_brms, ndraws = 25)
```

:::

## Posterior prediction

What if it's 10˚ at 9 AM in both cities? What's the predicted afternoon temperature?

::: {.panel-tabset}
### rstanarm

```{r}
weather_location_rstanarm |> 
  predicted_draws(newdata = expand_grid(temp9am = 10, 
                                        location = c("Wollongong", "Uluru"))) |> 
  ggplot(aes(x = .prediction, y = location, fill = location)) +
  stat_halfeye() +
  scale_fill_manual(values = c(clrs[2], clrs[1]), guide = "none") +
  labs(x = "Predicted 3 PM temperature", y = NULL)
```

### brms

```{r}
weather_location_brms |> 
  predicted_draws(newdata = expand_grid(temp9am_c = 10 - unscaled$temp9am_centered$scaled_center, 
                                        location = c("Wollongong", "Uluru"))) |> 
  ggplot(aes(x = .prediction, y = location, fill = location)) +
  stat_halfeye() +
  scale_fill_manual(values = c(clrs[2], clrs[1]), guide = "none") +
  labs(x = "Predicted 3 PM temperature", y = NULL)
```

:::


# 11.3: Utilizing interaction terms

Using location as an indicator variable made sense for 9 AM temperatures—the relationship was the same in both locations, but one was just shifted up compared to the other. But that's not always the case! Here's 9 AM humidity. It's wildly different across the two locations (again because Uluru is in the middle of the desert and Wollongong is on the coast).

```{r message=FALSE}
ggplot(weather_WU, aes(x = humidity9am, y = temp3pm, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "9 AM humidity", y = "3 PM temperature", color = NULL)
```

We can estimate a different slope for Wollongong by including an interaction term, which makes the model a little more complicated:

$$
\begin{aligned}
Y_i \mid \beta_0, \beta_1, \beta_2, \beta_3, \sigma &\stackrel{\text{ind}}{\sim} \mathcal{N}(\mu_i, \sigma^2) \\
\mu_i &= \beta_{0c} + \beta_1 X_{i1} + \beta_2 X_{i2} + \beta_3 (X_{i1} \times X_{i2}) \\
\\
\beta_{0c} &\sim \mathcal{N}(25, 5) \\
\beta_{1} &\sim \mathcal{N}(0, 2.5) \\
\beta_{2} &\sim \mathcal{N}(0, 10) \\
\beta_{3} &\sim \mathcal{N}(0, 2.5) \\
\sigma &\sim \operatorname{Exponential}(1)
\end{aligned}
$$

## Understanding the priors

These look fine I guess. Though only really for the autoscaled rstanarm priors. The super vague brms priors lead to some wild predictions. 

::: {.panel-tabset}
### rstanarm

```{r model-interaction-prior-rstanarm, cache=TRUE, results="hide"}
weather_interaction_prior_rstanarm <- stan_glm(
  temp3pm ~ humidity9am + location + humidity9am:location,
  data = weather_WU,
  family = gaussian,
  prior_intercept = normal(25, 5),
  prior = normal(0, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0,
  prior_PD = TRUE
)
```

```{r fig.width=8, fig.height=3.5}
p1 <- weather_WU |> 
  add_predicted_draws(weather_interaction_prior_rstanarm, ndraws = 100) |> 
  ggplot(aes(x = .prediction, group = .draw)) +
  geom_density(size = 0.25, color = clrs[3]) +
  labs(x = "Predicted 3 PM temperature", y = "Density")

# Feed add_epred_draws() brand new data instead of the original data, just for fun
prior_draws_rstanarm <- weather_WU |> 
  group_by(location) |> 
  summarize(min = min(humidity9am),
            max = max(humidity9am)) |> 
  mutate(humidity9am = map2(min, max, ~seq(.x, .y, 1))) |> 
  unnest(humidity9am) |> 
  add_epred_draws(weather_interaction_prior_rstanarm, ndraws = 100)

p2 <- prior_draws_rstanarm |> 
  ggplot(aes(x = humidity9am, y = .epred)) +
  geom_line(aes(group = paste(location, .draw), color = location), alpha = 0.2) +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "9 AM humidity", y = "3 PM temperature", color = NULL) +
  theme(legend.position = "bottom")

p1 | p2
```

### brms

```{r model-interaction-prior-brms, cache=TRUE, results="hide"}
priors <- c(prior(normal(25, 5), class = Intercept),
            prior(normal(0, 2.5), class = b, coef = "humidity9am_c"),
            prior(normal(0, 10), class = b, coef = "locationWollongong"),
            prior(normal(0, 2.5), class = b, coef = "humidity9am_c:locationWollongong"),
            prior(exponential(1), class = sigma))

weather_interaction_prior_brms <- brm(
  bf(temp3pm ~ humidity9am_c + location + humidity9am_c:location),
  data = weather_WU,
  family = gaussian(),
  prior = priors,
  chains = 4, iter = 5000*2, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0,
  sample_prior = "only"
)
```

```{r fig.width=8, fig.height=3.5}
p1 <- weather_WU |> 
  add_predicted_draws(weather_interaction_prior_brms, ndraws = 100) |> 
  ggplot(aes(x = .prediction, group = .draw)) +
  geom_density(size = 0.25, color = clrs[3]) +
  labs(x = "Predicted 3 PM temperature", y = "Density")

# Feed add_epred_draws() brand new data instead of the original data, just for fun
prior_draws_brms <- weather_WU |> 
  group_by(location) |> 
  summarize(min = min(humidity9am_c),
            max = max(humidity9am_c)) |> 
  mutate(humidity9am_c = map2(min, max, ~seq(.x, .y, 1))) |> 
  unnest(humidity9am_c) |> 
  add_epred_draws(weather_interaction_prior_brms, ndraws = 100) |> 
  mutate(humidity_9am = humidity9am_c + unscaled$humidity9am_centered$scaled_center)

p2 <- prior_draws_brms |> 
  ggplot(aes(x = humidity_9am, y = .epred)) +
  geom_line(aes(group = paste(location, .draw), color = location), alpha = 0.2) +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "9 AM humidity", y = "3 PM temperature", color = NULL) +
  theme(legend.position = "bottom")

p1 | p2
```

:::

## Simulating the posterior

::: {.panel-tabset}
### rstanarm

#### Model

```{r model-interaction-rstanarm, cache=TRUE, results="hide"}
weather_interaction_rstanarm <- stan_glm(
  temp3pm ~ humidity9am + location + humidity9am:location,
  data = weather_WU,
  family = gaussian,
  prior_intercept = normal(25, 5),
  prior = normal(0, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0
)
```

```{r}
tidy(weather_interaction_rstanarm, effects = c("fixed", "aux"), 
     conf.int = TRUE, conf.level = 0.8)
```

#### Coefficient posteriors

```{r}
weather_interaction_rstanarm |> 
  gather_draws(`(Intercept)`, humidity9am, locationWollongong, `humidity9am:locationWollongong`, sigma) |> 
  ungroup() |> 
  mutate(.variable = fct_relevel(factor(.variable), 
                                 c("(Intercept)", "humidity9am", "locationWollongong",
                                   "humidity9am:locationWollongong", "sigma"))) |> 
  ggplot(aes(x = .value, fill = .variable)) +
  stat_halfeye(normalize = "xy") +
  scale_fill_manual(values = c(clrs[1], clrs[3], clrs[4], clrs[2], clrs[5]), guide = "none") +
  facet_wrap(vars(.variable), scales = "free_x") +
  labs(x = "Parameter posterior") +
  theme(legend.position = "bottom")
```

#### Fitted draws

```{r}
weather_WU |> 
  add_linpred_draws(weather_interaction_rstanarm, ndraws = 100) |> 
  ggplot(aes(x = humidity9am, y = temp3pm, color = location)) +
  geom_point(data = weather_WU, size = 0.5) +
  geom_line(aes(y = .linpred, group = paste(location, .draw)), alpha = 0.2, size = 0.5) +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "9 AM humidity", y = "3 PM temperature", color = NULL)
```

#### Posterior predictive check

```{r}
pp_check(weather_interaction_rstanarm, n = 25)
```


### brms

#### Model

```{r model-interaction-brms, cache=TRUE, results="hide"}
priors <- c(prior(normal(25, 5), class = Intercept),
            prior(normal(0, 2.5), class = b, coef = "humidity9am_c"),
            prior(normal(0, 10), class = b, coef = "locationWollongong"),
            prior(normal(0, 2.5), class = b, coef = "humidity9am_c:locationWollongong"),
            prior(exponential(1), class = sigma))

weather_interaction_brms <- brm(
  bf(temp3pm ~ humidity9am_c + location + humidity9am_c:location),
  data = weather_WU,
  family = gaussian(),
  prior = priors,
  chains = 4, iter = 5000*2, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0
)
```

```{r}
tidy(weather_interaction_brms, conf.int = TRUE, conf.level = 0.8) |> 
  select(-c(effect, component, group))
```

#### Coefficient posteriors

```{r}
weather_interaction_brms |> 
  gather_draws(b_Intercept, b_humidity9am_c, b_locationWollongong, 
               `b_humidity9am_c:locationWollongong`, sigma) |> 
  ungroup() |> 
  mutate(.variable = fct_relevel(factor(.variable), 
                                 c("b_Intercept", "b_humidity9am_c", "b_locationWollongong", 
                                   "b_humidity9am_c:locationWollongong", "sigma"))) |> 
  ggplot(aes(x = .value, fill = .variable)) +
  stat_halfeye(normalize = "xy") +
  scale_fill_manual(values = c(clrs[1], clrs[3], clrs[4], clrs[2], clrs[5]), guide = "none") +
  facet_wrap(vars(.variable), scales = "free_x") +
  labs(x = "Parameter posterior") +
  theme(legend.position = "bottom")
```

#### Fitted draws

```{r}
weather_WU |> 
  add_linpred_draws(weather_interaction_brms, ndraws = 100) |> 
  ggplot(aes(x = humidity9am, y = temp3pm, color = location)) +
  geom_point(data = weather_WU, size = 0.5) +
  geom_line(aes(y = .linpred, group = paste(location, .draw)), alpha = 0.2, size = 0.5) +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  labs(x = "9 AM humidity", y = "3 PM temperature", color = NULL)
```

#### Posterior predictive check

```{r}
pp_check(weather_interaction_brms, ndraws = 25)
```

:::

## Posterior prediction

What if there's 50% humidity at 9 AM in both cities? What's the predicted afternoon temperature?

::: {.panel-tabset}
### rstanarm

```{r}
weather_interaction_rstanarm |> 
  predicted_draws(newdata = expand_grid(humidity9am = 50, 
                                        location = c("Wollongong", "Uluru"))) |> 
  ggplot(aes(x = .prediction, y = location, fill = location)) +
  stat_halfeye() +
  scale_fill_manual(values = c(clrs[2], clrs[1]), guide = "none") +
  labs(x = "Predicted 3 PM temperature", y = NULL)
```

### brms

```{r}
weather_interaction_brms |> 
  predicted_draws(newdata = expand_grid(humidity9am_c = 50 - unscaled$humidity9am_centered$scaled_center, 
                                        location = c("Wollongong", "Uluru"))) |> 
  ggplot(aes(x = .prediction, y = location, fill = location)) +
  stat_halfeye() +
  scale_fill_manual(values = c(clrs[2], clrs[1]), guide = "none") +
  labs(x = "Predicted 3 PM temperature", y = NULL)
```

:::


# 11.4: Utilizing more than 2 predictors

Let's go wild and use everything! 

$$
\begin{aligned}
Y_i \mid \beta_0, \beta_1, \beta_2, \beta_3, \beta_4, \beta_5, \sigma &\stackrel{\text{ind}}{\sim} \mathcal{N}(\mu_i, \sigma^2) \\
\mu_i &= \beta_{0c} + \beta_1 X_{i1} + \beta_2 X_{i2} + \beta_3 X_{i3} + \beta_4 X_{i4} + \beta_5 X_{i5}\\
\\
\beta_{0c} &\sim \mathcal{N}(25, 5) \\
\beta_{1} \dots \beta_5 &\sim \mathcal{N}(0, 2.5) \\
\sigma &\sim \operatorname{Exponential}(1)
\end{aligned}
$$

## Understanding the priors

These look fine to me. brms priors could be better, but whatever.

::: {.panel-tabset}
### rstanarm

```{r model-full-prior-rstanarm, cache=TRUE, results="hide"}
weather_full_prior_rstanarm <- stan_glm(
  temp3pm ~ temp9am + humidity9am + windspeed9am + pressure9am + location,
  data = weather_WU,
  family = gaussian,
  prior_intercept = normal(25, 5),
  prior = normal(0, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0,
  prior_PD = TRUE
)
```

```{r}
weather_WU |> 
  add_predicted_draws(weather_full_prior_rstanarm, ndraws = 100) |> 
  ggplot(aes(x = .prediction, group = .draw)) +
  geom_density(size = 0.25, color = clrs[3]) +
  labs(x = "Predicted 3 PM temperature", y = "Density")
```

### brms

```{r model-full-prior-brms, cache=TRUE, results="hide"}
priors <- c(prior(normal(25, 5), class = Intercept),
            prior(normal(0, 2.5), class = b),
            prior(exponential(1), class = sigma))

weather_full_prior_brms <- brm(
  bf(temp3pm ~ temp9am_c + humidity9am_c + windspeed9am_c + pressure9am_c + location),
  data = weather_WU,
  family = gaussian(),
  prior = priors,
  chains = 4, iter = 5000*2, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0,
  sample_prior = "only"
)
```

```{r}
weather_WU |> 
  add_predicted_draws(weather_full_prior_brms, ndraws = 100) |> 
  ggplot(aes(x = .prediction, group = .draw)) +
  geom_density(size = 0.25, color = clrs[3]) +
  labs(x = "Predicted 3 PM temperature", y = "Density")
```

:::

## Simulating the posterior

::: {.panel-tabset}
### rstanarm

#### Model

```{r model-full-rstanarm, cache=TRUE, results="hide"}
weather_full_rstanarm <- stan_glm(
  temp3pm ~ temp9am + humidity9am + windspeed9am + pressure9am + location,
  data = weather_WU,
  family = gaussian,
  prior_intercept = normal(25, 5),
  prior = normal(0, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0
)
```

```{r}
tidy(weather_full_rstanarm, effects = c("fixed", "aux"), 
     conf.int = TRUE, conf.level = 0.8)
```

#### Coefficient posteriors

```{r}
weather_full_rstanarm |> 
  gather_draws(`(Intercept)`, temp9am, humidity9am, windspeed9am, 
               pressure9am, locationWollongong, sigma) |> 
  ungroup() |> 
  mutate(.variable = fct_relevel(factor(.variable),
                                 c("(Intercept)", "temp9am", "humidity9am", "windspeed9am", 
                                   "pressure9am", "locationWollongong", "sigma"))) |> 
  ggplot(aes(x = .value, fill = .variable)) +
  stat_halfeye(normalize = "xy") +
  guides(fill = "none") +
  facet_wrap(vars(.variable), scales = "free_x") +
  labs(x = "Parameter posterior") +
  theme(legend.position = "bottom")
```

#### Posterior predictive check

Super nice now

```{r}
pp_check(weather_full_rstanarm, n = 25)
```


### brms

#### Model

```{r model-full-brms, cache=TRUE, results="hide"}
priors <- c(prior(normal(25, 5), class = Intercept),
            prior(normal(0, 2.5), class = b),
            prior(exponential(1), class = sigma))

weather_full_brms <- brm(
  bf(temp3pm ~ temp9am_c + humidity9am_c + windspeed9am_c + pressure9am_c + location),
  data = weather_WU,
  family = gaussian(),
  prior = priors,
  chains = 4, iter = 5000*2, seed = BAYES_SEED, 
  backend = "cmdstanr", refresh = 0
)
```

```{r}
tidy(weather_full_brms, conf.int = TRUE, conf.level = 0.8) |> 
  select(-c(effect, component, group))
```

#### Coefficient posteriors

```{r}
get_variables(weather_full_brms)
weather_full_brms |> 
  gather_draws(b_Intercept, b_temp9am_c, b_humidity9am_c, b_windspeed9am_c,
               b_pressure9am_c, b_locationWollongong, sigma) |> 
  ungroup() |> 
  mutate(.variable = fct_relevel(factor(.variable), 
                                 c("b_Intercept", "b_temp9am_c", "b_humidity9am_c", "b_windspeed9am_c",
                                   "b_pressure9am_c", "b_locationWollongong", "sigma"))) |> 
  ggplot(aes(x = .value, fill = .variable)) +
  stat_halfeye(normalize = "xy") +
  guides(fill = "none") +
  facet_wrap(vars(.variable), scales = "free_x") +
  labs(x = "Parameter posterior") +
  theme(legend.position = "bottom")
```

#### Posterior predictive check

Delightful.

```{r}
pp_check(weather_full_brms, ndraws = 25)
```

:::

## Posterior prediction

What if there's 50% humidity, 10 MPH of wind, pressure of 1015, and temperature of 10˚ at 9 AM in both cities? What's the predicted afternoon temperature?

::: {.panel-tabset}
### rstanarm

```{r}
weather_full_rstanarm |> 
  predicted_draws(newdata = expand_grid(humidity9am = 50, 
                                        windspeed9am = 10,
                                        pressure9am = 1015,
                                        temp9am = 10,
                                        location = c("Wollongong", "Uluru"))) |> 
  ggplot(aes(x = .prediction, y = location, fill = location)) +
  stat_halfeye() +
  scale_fill_manual(values = c(clrs[2], clrs[1]), guide = "none") +
  labs(x = "Predicted 3 PM temperature", y = NULL)
```

### brms

```{r}
weather_full_brms |> 
  predicted_draws(newdata = expand_grid(humidity9am_c = 50 - unscaled$humidity9am_centered$scaled_center, 
                                        windspeed9am_c = 10 - unscaled$windspeed9am_centered$scaled_center,
                                        pressure9am_c = 1015 - unscaled$pressure9am_centered$scaled_center,
                                        temp9am_c = 10 - unscaled$temp9am_centered$scaled_center,
                                        location = c("Wollongong", "Uluru"))) |> 
  ggplot(aes(x = .prediction, y = location, fill = location)) +
  stat_halfeye() +
  scale_fill_manual(values = c(clrs[2], clrs[1]), guide = "none") +
  labs(x = "Predicted 3 PM temperature", y = NULL)
```

:::


# 11.5: Model evaluation and comparison

Phew. We have 5 (actually 10) different models predicting afternoon temperature:

+-------------------------------------------------------------------+--------------------------------------------+
| `weather_simple_rstanarm` and `weather_simple_brms`               |     temp3pm ~ temp9am                      |
+-------------------------------------------------------------------+--------------------------------------------+
| `weather_location_only_rstanarm` and `weather_location_only_brms` |     temp3pm ~ location                     |
+-------------------------------------------------------------------+--------------------------------------------+
| `weather_location_rstanarm` and `weather_location_brms`           |     temp3pm ~ temp9am + location           |
+-------------------------------------------------------------------+--------------------------------------------+
| `weather_interaction_rstanarm` and `weather_interaction_brms`     |     temp3pm ~ humidity9am + location +         |
|                                                                   |               humidity9am:location             |
+-------------------------------------------------------------------+--------------------------------------------+
| `weather_full_rstanarm` and `weather_full_brms`                   |     temp3pm ~ temp9am + humidity9am +      |
|                                                                   |               windspeed9am + pressure9am + |
|                                                                   |               location                     |
+-------------------------------------------------------------------+--------------------------------------------+


## How fair is each model? 

That's fine here. They're all equally fair (and the data was collected through an automated system and seems fair)

## How wrong is each model?

We can compare all the posterior predictive checks. `y_rep` gets better as the model gets more complex.

::: {.panel-tabset}
### rstanarm

```{r post-predict-rstanarm, cache=TRUE}
models_rstanarm <- tribble(
  ~model_name, ~model,
  "Temperature only", weather_simple_rstanarm, 
  "Location only", weather_location_only_rstanarm, 
  "Temperature + location", weather_location_rstanarm, 
  "Temperature × location", weather_interaction_rstanarm, 
  "Full model", weather_full_rstanarm
) |> 
  mutate(ppcheck = map2(model, model_name, ~pp_check(.x, n = 25) + labs(title = .y))) |> 
  mutate(loo = map(model, ~loo(.)))
```


```{r fig.width=8, fig.height=4.5}
wrap_plots(models_rstanarm$ppcheck)
```

### brms

```{r post-predict-brms, cache=TRUE, warning=FALSE}
models_brms <- tribble(
  ~model_name, ~model,
  "Temperature only", weather_simple_brms, 
  "Location only", weather_location_only_brms, 
  "Temperature + location", weather_location_brms, 
  "Temperature × location", weather_interaction_brms, 
  "Full model", weather_full_brms
) |> 
  mutate(ppcheck = map2(model, model_name, ~pp_check(.x, ndraws = 25) + labs(title = .y))) |> 
  mutate(loo = map(model, ~loo(.)))
```

```{r fig.width=8, fig.height=4.5}
wrap_plots(models_brms$ppcheck)
```

:::

## How accurate are each model's predictions?

We can check these with the 3 approaches from chapter 10.

### Visualization

For the models with fewer dimensions we can visualize their predictions, like with the location-only model and the model with the interaction

::: {.panel-tabset}
#### rstanarm

```{r}
weather_location_only_rstanarm |> 
  add_predicted_draws(newdata = weather_WU, ndraws = 100) |> 
  ggplot(aes(x = location)) +
  stat_dotsinterval(aes(y = .prediction, color = location, fill = location), 
                    quantiles = 300, side = "right", slab_size = 0, scale = 0.4) +
  stat_dotsinterval(aes(y = temp3pm, fill = location),
                    side = "left", quantiles = 100, slab_size = 0, slab_shape = 23) +
  scale_fill_manual(values = c(clrs[2], clrs[1]), guide = "none") +
  guides(color = "none") +
  labs(title = "Actual data on left; predictions on right",
       y = "Predicted afternoon temperature", x = NULL)
```

#### brms

```{r}
weather_location_only_brms |> 
  add_predicted_draws(newdata = weather_WU, ndraws = 100) |> 
  ggplot(aes(x = location)) +
  stat_dotsinterval(aes(y = .prediction, color = location, fill = location), 
                    quantiles = 300, side = "right", slab_size = 0, scale = 0.4) +
  stat_dotsinterval(aes(y = temp3pm, fill = location),
                    side = "left", quantiles = 100, slab_size = 0, slab_shape = 23) +
  scale_fill_manual(values = c(clrs[2], clrs[1]), guide = "none") +
  guides(color = "none") +
  labs(title = "Actual data on left; predictions on right",
       y = "Predicted afternoon temperature", x = NULL)
```
:::

::: {.panel-tabset}
#### rstanarm

```{r}
weather_interaction_rstanarm |> 
  add_predicted_draws(newdata = weather_WU, ndraws = 100) |> 
  ggplot(aes(x = temp9am)) +
  stat_interval(aes(y = .prediction, color = location, color_ramp = stat(level)), alpha = 0.25,
                .width = c(0.5, 0.89, 0.95)) +
  geom_point(aes(y = temp3pm, fill = location), size = 2, pch = 21, color = "white") +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  scale_fill_manual(values = c(clrs[2], clrs[1]))
```

#### brms

```{r}
weather_interaction_brms |> 
  add_predicted_draws(newdata = weather_WU, ndraws = 100) |> 
  ggplot(aes(x = temp9am)) +
  stat_interval(aes(y = .prediction, color = location, color_ramp = stat(level)), alpha = 0.25,
                .width = c(0.5, 0.89, 0.95)) +
  geom_point(aes(y = temp3pm, fill = location), size = 2, pch = 21, color = "white") +
  scale_color_manual(values = c(clrs[2], clrs[1])) +
  scale_fill_manual(values = c(clrs[2], clrs[1]))
```
:::

We can't visualize the predictions from the full model without holding a bunch of stuff constant, though. So we use numbers instead!

### Cross validation

We could run k-fold cross validation here, but we'd need to run 100 different Stan models (10 folds for each of the 10 models) and I don't want to do that. The book has results from the rstanarm models. It shows that the smaller `temp9am + location` model is the most efficient of the bunch.

### ELPD

The full model and temperature + location models have the highest ELPD, but they're not significantly different from each other, so they perform about the same. The Bayes Rules! authors conclude:

> Again, since [the temperature + location model] is simpler and achieves similar predictive accuracy, we’d personally choose to use it over the more complicated [full model].

::: {.panel-tabset}
#### rstanarm

```{r}
loo_rstanarm <- models_rstanarm |> 
  mutate(loo_stuff = map(loo, ~as_tibble(.$estimates, rownames = "statistic"))) |> 
  select(model_name, loo_stuff) |> 
  unnest(loo_stuff) |> 
  filter(statistic == "elpd_loo") |> 
  arrange(desc(Estimate))
loo_rstanarm
```

```{r}
loo_rstanarm |> 
  mutate(model_name = fct_rev(fct_inorder(model_name))) |> 
  ggplot(aes(x = Estimate, y = model_name)) +
  geom_pointrange(aes(xmin = Estimate - 2 * SE, xmax = Estimate + 2 * SE))
```

```{r}
models_rstanarm |> 
  pull(loo) |> 
  set_names(models_rstanarm$model_name) |> 
  loo_compare()
```

#### brms
```{r}
loo_brms <- models_brms |> 
  mutate(loo_stuff = map(loo, ~as_tibble(.$estimates, rownames = "statistic"))) |> 
  select(model_name, loo_stuff) |> 
  unnest(loo_stuff) |> 
  filter(statistic == "elpd_loo") |> 
  arrange(desc(Estimate))
loo_brms
```

```{r}
loo_brms |> 
  mutate(model_name = fct_rev(fct_inorder(model_name))) |> 
  ggplot(aes(x = Estimate, y = model_name)) +
  geom_pointrange(aes(xmin = Estimate - 2 * SE, xmax = Estimate + 2 * SE))
```

```{r}
models_brms |> 
  pull(loo) |> 
  set_names(models_brms$model_name) |> 
  loo_compare()
```
:::
